---
title: "Assignment 5: Data Visualization"
author: "Priya Rathod"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
fig_height: 3
fig_width: 4
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Visualization 

## Directions
1. Rename this file `<FirstLast>_A05_DataVisualization.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.

---

## Set up your session 

1. Set up your session. Load the tidyverse, lubridate, here & cowplot packages, and verify your home directory. Read in the NTL-LTER processed data files for nutrients and chemistry/physics for Peter and Paul Lakes (use the tidy `NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv` version in the Processed_KEY folder) and the processed data file for the Niwot Ridge litter dataset (use the `NEON_NIWO_Litter_mass_trap_Processed.csv` version, again from the Processed_KEY folder). 

2. Make sure R is reading dates as date format; if not change the format to date.

```{r initialize project}
#1 

library(ggplot2)
library(tidyverse)
library(lubridate)
library(here)
library(cowplot)
here()


lake_chem_nutrients <- 
  read.csv(here("Data/Processed/Processed_KEY/Processed_KEY/NTL-LTER_Lake_Chemistry_Nutrients_PeterPaul_Processed.csv"), stringsAsFactors = TRUE)

neon_litter <- 
  read.csv(here("Data/Processed/Processed_KEY/Processed_KEY/NEON_NIWO_Litter_mass_trap_Processed.csv"), stringsAsFactors = TRUE)


#2 

lake_chem_nutrients$sampledate <- ymd(lake_chem_nutrients$sampledate)
class(lake_chem_nutrients$sampledate)

neon_litter$collectDate <- ymd(neon_litter$collectDate)
class(neon_litter$collectDate)


```
## Define your theme
3. Build a theme and set it as your default theme. Customize the look of at least two of the following: 
 * Plot background
 * Plot title
 * Axis labels
 * Axis ticks/gridlines
 * Legend

```{r build plot theme}
#3


mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.direction = "vertical", plot.background = element_rect(fill = "white")) 

theme_set(mytheme)

```

## Create graphs
For numbers 4-7, create ggplot graphs and adjust aesthetics to follow best practices for data visualization. Ensure your theme, color palettes, axes, and additional aesthetics are edited accordingly.

4. [NTL-LTER] Plot total phosphorus (`tp_ug`) by phosphate (`po4`), with separate aesthetics for Peter and Paul lakes. Add line(s) of best fit using the `lm` method. Adjust your axes to hide extreme values (hint: change the limits using `xlim()` and/or `ylim()`).

```{r plot total P vs PO4}
#4 

plot_nutrients <- ggplot(lake_chem_nutrients, aes(x = tp_ug, y = po4, color = lakename)) +
  xlim(0,150) + ylim(0,50) + geom_smooth(method=lm) + 
  labs(title = "Phosphorus and Nitrogen - Peter and Paul Lakes",
x = "Phosphorus in mg/L", y = "Nitrogen in mg/L") + geom_point() + scale_color_manual(values = c("#7fcdbb", "#225ea8"))
print(plot_nutrients)


```


5. [NTL-LTER] Make three separate boxplots of (a) temperature, (b) TP, and (c) TN, with month as the x axis and lake as a color aesthetic. Then, create a cowplot that combines the three graphs. Make sure that only one legend is present and that graph axes are aligned. 

Tips:
* Recall the discussion on factors in the lab section as it may be helpful here. 
* Setting an axis title in your theme to `element_blank()` removes the axis title (useful when multiple, aligned plots use the same axis values)
* Setting a legend's position to "none" will remove the legend from a plot.
* Individual plots can have different sizes when combined using `cowplot`.

```{r Create boxplots}
#5 

#Converting month to factors
lake_chem_nutrients$month <- factor(lake_chem_nutrients$month)

#Creating list of months for axis labels
month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", 
           "Oct", "Nov", "Dec")

#Plotting Temperature
plot_temp <-
  ggplot(lake_chem_nutrients, aes(x = month, y = temperature_C, 
 color = lakename)) +
  geom_boxplot() + scale_x_discrete(limits = as.character(1:12), labels =  c(month)) + mytheme +
  labs(title = "Monthly Temperature - Peter and Paul Lakes", 
       x = "Month", y = "Temperature in °C") + 
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + 
  scale_color_manual(values = c("#2246be", "#1d9a0c"))
print(plot_temp)

#Plotting Phosphorus
plot_tp <-
  ggplot(lake_chem_nutrients, aes(x = month, y = tp_ug,
 color = lakename)) +
  geom_boxplot() + scale_x_discrete(limits = as.character(1:12), labels =  c(month)) + mytheme +
labs(title = "Total Phosphorus by Month - Peter and Paul Lakes",
     x = "Month", y = "Phosphorus in mg/L") +
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + 
  scale_color_manual(values = c("#2246be", "#1d9a0c"))
print(plot_tp)

#Plotting Nitrogen
plot_tn <-
  ggplot(lake_chem_nutrients, aes(x = month, y = tn_ug,
color = lakename)) +
  geom_boxplot() + scale_x_discrete(limits = as.character(1:12), 
        labels =  c(month)) + mytheme +
  labs(title = "Total Nitrogen by Month - Peter and Paul Lakes", x = "Month",
       y = "Nitrogen in mg/L") + theme(legend.position = "right") + theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + 
  scale_color_manual(values = c("#2246be", "#1d9a0c"))
print(plot_tn)

#Creating clean plots - without labels

plot_temp_clean <- plot_temp + labs(title = NULL, x = NULL) + theme(legend.position = "none") + theme(plot.margin = unit(c(1,0,1,0), "cm"))

plot_tp_clean <- plot_tp + labs(title = NULL, x = NULL) +
theme(legend.position = "none") + theme(plot.margin = unit(c(1,0,1,0), "cm"))

plot_tn_clean <- plot_tn + labs(title = NULL, x = NULL) +
theme(legend.position = "none") + theme(plot.margin = unit(c(1,0,1,0), "cm"))

#Extracting legend

legend <- get_legend(plot_temp_clean + theme(legend.position = "right"))

#Creating the cowplot

combined_plot <- plot_grid(plot_temp_clean, plot_tp_clean, plot_tn_clean,
     legend, align = 'h', nrow = 1, rel_widths = c(1,1,1,0.5)) +
draw_label("Temperature, Phosphorus and Nitrogen Levels by Month - Peter and Paul Lakes", x = 0.5, y = 0.98) + draw_label("Month", x = 0.5, y = 0.04) + scale_color_manual(values = c("#2246be", "#1d9a0c"))
print(combined_plot)

```

Question: What do you observe about the variables of interest over seasons and between lakes? 

> Answer: Overall, there seems to be a trend with increasing levels of
phosphorus and nitrogen as the temperature rises - with levels peaking roughly
between August and September.

6. [Niwot Ridge] Plot a subset of the litter dataset by displaying only the "Needles" functional group. Plot the dry mass of needle litter by date and separate by NLCD class with a color aesthetic. (no need to adjust the name of each land use)

7. [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

```{r Plot litter}
#6

#Converting month to factors
neon_litter$collectDate <- factor(neon_litter$collectDate)

#Separating needle litter data and grouping date by month and year 

needle_litter <- neon_litter  %>% 
  filter(functionalGroup == "Needles") 

needle_litter <- needle_litter %>%
  mutate(year = year(collectDate),
    month = month(collectDate))

plot_needle <-
  ggplot(needle_litter, aes(x = year, y = dryMass,
 color = nlcdClass)) + geom_point() + mytheme +
  labs(title = "Annual Needle Litter Mass (2016-2019)", x = "Year",
       y = "Dry mass in Kg", color = "Land use type") + 
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + 
  scale_color_manual(values = c("#ea5d0d", "#1d9a0c", "#600c9a"))
print(plot_needle)
  
#7 [Niwot Ridge] Now, plot the same plot but with NLCD classes separated into three facets rather than separated by color. 

plot_needle_facets <-
  ggplot(needle_litter, aes(x = year, y = dryMass)) +
  geom_point(size=1) + mytheme + labs(title = "Annual Needle Litter Mass (2016-2019)", x = "Year", y = "Dry mass in Kg", color = "Land use type") + theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) +
  facet_wrap(vars(nlcdClass), nrow = 3)
print(plot_needle_facets)

plot_needle_facets_colors <-
  ggplot(needle_litter, aes(x = year, y = dryMass, color = nlcdClass)) + geom_point(size=1) + mytheme + labs(title = "Annual Needle Litter Mass (2016-2019)", 
x = "Year", y = "Dry mass in Kg", color = "Land use type") +
theme(legend.position = "right") + 
theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + facet_wrap(vars(nlcdClass), nrow = 3) + scale_color_manual(values = c("#ea5d0d", "#1d9a0c", "#600c9a"))
print(plot_needle_facets_colors)


```
Question: Which of these plots (6 vs. 7) do you think is more effective, and why? 

> Answer: I think that the plot produced for question 7 (facets) or the 
additional graph that I made (facets and colors) is more effective.
Within each land use type, the data is largely concentrated between 0-20 Kgs.
With three colors for each land use type, it is difficult to differentiate.
With the facets and the colors, it is easier to identify the annual trends and 
differences between land use type, since the data points have been separated out.