---
title: "Assignment 3: Data Exploration"
author: "Priya Rathod"
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}

#library(tidyverse) ##loading packages
#library(lubridate)
#library(here)
#library(dplyr)

getwd() ## checking working directory

neonics <- read.csv(
  file = here("Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"),
  stringsAsFactors = TRUE)

litter <- read.csv(
  file = here("Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE)

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Neonicotinoids are known to disturb some critical functions that
insects rely on plants for (e.g.: pollination). We would be interested in 
understanding how insect populations are impacted by the increasing use of
neonicotinoids.


3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer: Organic litter and woody debris form important ecosystems on the
forest floor. These dead trees support nutrient recycling, microhabitats for 
wildlife and soil stability. These topics can provide important insights into
the health of these forest ecosystems.

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1. Sampling occured at sites that contain woody vegetation that was greater
 than 2m.
 2.Locations of the sampling plots were chosen randomly.
 3. Sampling frequency varied based on ground and elevated traps and vegetation
 type.

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}

dim(neonics)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}

summary(neonics$Effect)
sort(summary(neonics$Effect))

```

> Answer: The most common effects studied are population and mortality effects.
These would be of specific interest since the neonicotinoids are known to
disturb plant function, thereby impacting insect birth and death rates.


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}

#View(neonics)
summary((neonics$Species.Common.Name), maxsum = 7)

```

> Answer: The six most commonly studied species are all bee/wasp species, 
part of the Hymenoptera order. They are all important pollinators and
hence, would be of interest in understanding how the neonicotinoids' use
impacts insect populations.

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}

#View(neonics)
neonics$Conc.1..Author.
class(neonics$Conc.1..Author.)

```

> Answer: The class of `Conc.1..Author is a factor. It is not numeric because
the data includes characters like "~", "<", "/".

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}

ggplot(neonics) +
  geom_freqpoly(aes(x = Publication.Year), binwidth = 1) +  
  labs(title = "Number of studies conducted between 1980-2020", 
                     x = "Publication Year", y = "Number of studies")

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}

ggplot(neonics) +
  geom_freqpoly(aes(x = Publication.Year, color = Test.Location), binwidth = 1) +  
  labs(title = "Number of studies conducted between 1980-2020", 
                     x = "Publication Year", y = "Number of studies")

```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: The most common test locations are the 'lab' and 'field natural'
with studies with lab locations growing between 2000-2020 and studies with
'field natural' locations declining in the same time period.

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}

ggplot(data = neonics, aes(x = Endpoint)) +
  geom_bar() + theme(axis.text.x = element_text(angle = 90,
                                                vjust = 0.5, hjust=1)) + 
  labs(title = "Number of studies conducted between 1980-2020", 
                     x = "Publication Year", y = "Number of studies")

```

> Answer: The two most common end points are LOEL
(Lowest-observable-effect-level) and NOEL
(No-observable-effect-level).

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}

class(litter$collectDate)
#View(litter$collectDate)
litter_new_date <- as.Date(litter$collectDate)
class(litter_new_date) #new values are dates

unique(litter_new_date) #Data was collected on two days in August - 2nd and 30th.

```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}

#View(litter)
#View(unique(litter$plotID)) # 12 plots
summary(litter$plotID)
```

> Answer: 'summary' shows us a count of rows per plot ID whereas 'unique' lists only the unique values within the column.

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}

ggplot(data = litter, aes(x = litter$functionalGroup)) +
  geom_bar() + labs(title = "Litter by Functional Group", x = "Functional groups", y = "Count observed")

```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}

ggplot(litter) +
  geom_boxplot(aes(x = functionalGroup, y = dryMass)) + labs(title = "Weight of Litter by Functional Group", x = "Functional groups", y = "Dry Mass")

ggplot(litter) +
  geom_violin(aes(x = functionalGroup, y = dryMass)) + labs(title = "Weight of Litter by Functional Group", x = "Functional groups", y = "Dry Mass")


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer: We have very few datapoints per functional group in this data set (20-30 per group). There is also very little range within the functional groups, as the box plot shows.

What type(s) of litter tend to have the highest biomass at these sites?
```{r}
litter %>%
+     group_by(functionalGroup) %>%
+     summarize(dryMass = n())

#Double checking with a pivot table to see the data
```

> Answer: Needles tend to have the highest biomasss at these sites.
