---
title: "Assignment 10: Data Scraping"
author: "Priya Rathod"
output: pdf_document
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on data scraping. 

## Directions
1. Rename this file `<FirstLast>_A10_DataScraping.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure your code is tidy; use line breaks to ensure your code fits in the knitted output.
5. Be sure to **answer the questions** in this assignment document.
6. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up 
1. Set up your session:

* Load the packages `tidyverse`, `rvest`, and any others you end up using.
* Check your working directory

```{r 1. Project Setup, message = FALSE}
#1 
library(tidyverse)
library(lubridate); library(dplyr)
library(here); here()

#install.packages("rvest")
library(rvest)

mytheme <- theme_classic(base_size = 14) +
  theme(axis.text = element_text(color = "black"), 
        legend.direction = "vertical", plot.background = element_rect(fill = "white")) 

theme_set(mytheme)
```

2. We will be scraping data from the NC DEQs Local Water Supply Planning website, specifically the Durham's 2023 Municipal Local Water Supply Plan (LWSP): 
 * Navigate to https://www.ncwater.org/WUDC/app/LWSP/search.php
 * Scroll down and select the LWSP link next to Durham Municipality. 
 * Note the web address: <https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023>
 
Indicate this website as the as the URL to be scraped. (In other words, read the contents into an `rvest` webpage object.)

```{r set.the.scraping.website}
#2 

webpage <- read_html('https://www.ncwater.org/WUDC/app/LWSP/report.php?pwsid=03-32-010&year=2023')

webpage

```

3. The data we want to collect are listed below:

* From the "1. System Information" section:
 * Water system name
 * PWSID
 * Ownership
 
* From the "3. Water Supply Sources" section:
 * Maximum Day Use (MGD) - for each month

In the code chunk below scrape these values, assigning them to four separate variables.

>HINT: The first value should be "Durham", the second "03-32-010", the third "Municipality", and the last should be a vector of 12 numeric values (represented as strings)".

```{r scrape.the.data}
#3 
water_system_name <- webpage %>% 
  html_nodes("div+ table tr:nth-child(1) td:nth-child(2)") %>% 
  html_text()
water_system_name

pwsid <- webpage %>% 
  html_nodes("td tr:nth-child(1) td:nth-child(5)") %>% 
  html_text()
pwsid

ownership <- webpage %>% 
  html_nodes("div+ table tr:nth-child(2) td:nth-child(4)") %>% 
  html_text()
ownership

max_daily_use <- webpage %>% 
  html_nodes("th~ td+ td") %>% 
  html_text()
max_daily_use

```


4. Convert your scraped data into a dataframe. This dataframe should have a column for each of the 4 variables scraped and a row for the month corresponding to the withdrawal data. Also add a Date column that includes your month and year in data format. (Feel free to add a Year column too, if you wish.)

>TIP: Use `rep()` to repeat a value when creating a dataframe.

>NOTE: It's likely you won't be able to scrape the monthly widthrawal data in chronological order. You can overcome this by creating a month column manually assigning values in the order the data are scraped: "Jan", "May", "Sept", "Feb", etc... Or, you could scrape month values from the web page...

5. Create a line plot of the maximum daily withdrawals across the months for 2023, making sure, the months are presented in proper sequence.

```{r create.a.dataframe.from.scraped.data}
#4 

df_withdrawals <- data.frame("Month" = rep(1:12),
                             "Year" = rep(2023,12),
                            "Max_daily_use" = as.numeric(max_daily_use)) 

month <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", 
           "Oct", "Nov", "Dec")

water_df <- df_withdrawals %>% 
  mutate(Water_System_Name = !!water_system_name,
         PWSID = !!pwsid,
         Ownership = !!ownership,
         Date = my(paste(Month,"-",Year)))

#5 

plot_water <-
  ggplot(water_df, aes(x = Date, y = Max_daily_use)) +
      geom_line() +  geom_smooth(method="loess",se=FALSE) +
 mytheme +
  labs(title = "Water Usage Data in Durham, NC in 2023", 
       x = "Month", y = "Maximum Daily Withdrawal, mgd") + 
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + 
 scale_x_date(date_breaks = "1 month", date_labels = "%b %Y")
print(plot_water)


```

6. Note that the PWSID and the year appear in the web address for the page we scraped. Construct a function using your code above that can scrape data for any PWSID and year for which the NC DEQ has data, returning a dataframe.
**Be sure to modify the code to reflect the year and site (pwsid) scraped**.

```{r construct.a.scraping.function}
#6. 
the_base_url <- 'https://www.ncwater.org/WUDC/app/LWSP/report.php?'
the_pwsid <- '03-32-010'
the_year <- 2014
the_scraped_url <- paste0(the_base_url, 'pwsid=', the_pwsid, '&year=', the_year)
print(the_scraped_url)

#Function 

scrape.it <- function(the_year, the_pwsid){
  
  the_website <- read_html(paste0('https://www.ncwater.org/WUDC/app/LWSP/report.php?', 'pwsid=', the_pwsid, '&year=', the_year))
  
the_system_tag <- 'div+ table tr:nth-child(1) td:nth-child(2)'
the_ownership_tag <- 'div+ table tr:nth-child(2) td:nth-child(4)'
the_pwsid_tag <- 'td tr:nth-child(1) td:nth-child(5)'
the_mgd_tag <- 'th~ td+ td'
  
the_system <- the_website %>% html_nodes(the_system_tag) %>% html_text()
the_ownership <- the_website %>% html_nodes(the_ownership_tag) %>% html_text()
the_pwsid <- the_website %>% html_nodes(the_pwsid_tag) %>% html_text()
the_mgd <- the_website %>% html_nodes(the_mgd_tag) %>% html_text()
  
df_withdrawals <- data.frame("Month" = rep(1:12),
 "Year" = rep(the_year,12),
 "Max_Daily_Withdrawals" = as.numeric(the_mgd)) %>% 
mutate(System_Name = !!the_system,
Ownership = !!the_ownership,
PWSID = !!the_pwsid,
Date = my(paste(Month,"-",Year)))

return(df_withdrawals)
}

```

7. Use the function above to extract and plot max daily withdrawals for Durham (PWSID='03-32-010') for each month in 2015

```{r fetch.and.plot.Durham.2015.data}
#7 

durham_2015 <- scrape.it(2015,'03-32-010')
view(durham_2015)

plot_water_durham <-
  ggplot(durham_2015, aes(x = Date, y = Max_Daily_Withdrawals)) +
      geom_line() +  geom_smooth(method="loess",se=FALSE) +
 mytheme +
  labs(title = "Water Usage Data in Durham, NC in 2015", 
       x = "Month", y = "Maximum Daily Withdrawal, mgd") + 
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + 
 scale_x_date(date_breaks = "1 month", date_labels = "%b %Y")
print(plot_water_durham)

```

8. Use the function above to extract data for Asheville (PWSID = 01-11-010) in 2015. Combine this data with the Durham data collected above and create a plot that compares Asheville's to Durham's water withdrawals.

```{r fetch.and.plot.Asheville.2015.data}
#8 

asheville_2015 <- scrape.it(2015,'01-11-010')

durham_ash_mgd_2015 <- rbind(durham_2015, asheville_2015)

plot_durham_ash_mgd <-
  ggplot(durham_ash_mgd_2015, aes(x = Date, y = Max_Daily_Withdrawals, color=System_Name)) +
      geom_line() +  geom_smooth(method="loess",se=FALSE) +
 mytheme +
  labs(title = "Water Usage Data in Durham and Asheville, NC in 2015", 
       x = "Month", y = "Maximum Daily Withdrawal, mgd") + 
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + 
 scale_x_date(date_breaks = "1 month", date_labels = "%b %Y")
print(plot_durham_ash_mgd)

```


9. Use the code & function you created above to plot Asheville's max daily withdrawal by months for the years 2018 thru 2022.Add a smoothed line to the plot (method = 'loess').

> TIP: See Section 3.2 in the "10_Data_Scraping.Rmd" where we apply "map2()" to iteratively run a function over two inputs. Pipe the output of the map2() function to `bindrows()` to combine the dataframes into a single one. 

```{r fetch.and.plot.Asheville.multiyear.data }
#9 

years = rep(2018:2022)
asheville_pwsid <- '01-11-010'
asheville_mgd <- map2(years, asheville_pwsid, scrape.it) %>% bind_rows()

plot_asheville_2022 <-
  ggplot(asheville_mgd, aes(x = Date, y = Max_Daily_Withdrawals)) +
      geom_line() +  geom_smooth(method="loess",se=FALSE) +
 mytheme +
  labs(title = "Water Usage Data in Asheville, NC between 2018-2022", 
       x = "Month", y = "Maximum Daily Withdrawal, mgd") + 
  theme(legend.position = "right") + 
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) + theme(axis.text=element_text(size=8)) + 
 scale_x_date(date_breaks = "4 month", date_labels = "%b %Y")
print(plot_asheville_2022)

```

Question: Just by looking at the plot (i.e. not running statistics), does Asheville have a trend in water usage over time?
> Answer: Yes. Overall, Asheville's maximum daily water use is increasing
between 2018-220. There was a slight decrease, in early 2020, after which daily water use has been increasing gradually. 
>
